Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations

### 2015-08-08 docs/response_to_reviewers.md

## Response to Reviewers

### Reviewer 1

We thank Reviewer 1 for the helpful suggestions. We agree with Reviewer 1 that
host-parasite distributions are dynamic and that it would be interesting to
apply the methods developed in this paper to a stochastic host-parasite model. However, we believe that this approach is unnecessary/problematic to include in the current manuscript for a few reasons.

First, stochastic host-parasite models can range from relatively simple models
with an analytical solutions at any given point in time (e.g. a simple death-
immigration process) to complex models with clumped infections, host
heterogeneity, and acquired immunity that can only be explored via simulation.
Given the space alloted in this manuscript and the time alloted for these
revisions, in would be difficult to provide a comprehensive treatment of how our
method works across the numerous different assumptions that have been included
in stochastic host-parasite models. We believe that this exploration would be
better suited for a follow-up study.

However, we can gain some insight into whether our model might be successful in
identifying PIHM in a simple birth-death-immigration process in which the number
of parasites increase by one, decrease by one, or stay the same in a given time
step.  The stationary distribution for a BDI process in which birth rate < death
rate and there is no parasite-induced host mortality is a negative binomial
distribution with $k =$ immigration rate / birth rate. This result is consistent
with our assumption that the pre-mortality distribution of parasites across
hosts is negative binomial. If we were to include parasite-induced host
mortality into this stochastic process, we know from extensive literature that
the variance to mean ratio would decrease and the parasite distribution with
PIHM would be less aggregated than the distribution without PIHM.  If the
immigration and birth rate (or self-reinfection rate) was known for the system,
our model could be applied in which the null model was a negative binomial with
a fixed $k = $ immigration / birth and the alternative was model was defined by
equation X in the manuscript.  If the immigration and birth rate was not known, our method could be applied as given in the manuscript.  The key criteria necessary for our method to work would be that the resulting distribution post mortality took a form significantly different than a negative binomial distribution. We have included an additional Appendix (Appendix X) in which we simulate a birth-death-immigration process with host mortality and test the applicability of our method.  We have also added the following the the Discussion to address Reviewer 1's comments.

@@

### Reviewer 2

We thank Reviewer 2 for their helpful feedback and address each of their points in turn.

1. Normally I would be asking authors to insert caveats into conclusions when they over spin their work, but in your case I do think you are way too negative about your own work when you talk about the caveats in the discussion and your conclusion / statements in the abstract.

We thank Reviewer 2 for this comment and have made changes in the Abstract and Discussion to cast our method in a more positive light.  In particular, we changed the the following sentences in the abstract from

*However, we stress that this method, and all techniques that estimate parasite- induced mortality from intensity data alone, have several critical assumptions that limit their applicability in the real world. Due to these limitations, these methods should only be used as an exploratory tool to inform more rigorous studies of parasite-induced host mortality.*

to

**However, this method, and all techniques that estimate parasite-induced
mortality from intensity data alone, have several important assumptions that
must be scrutinized before applying them to real-world data. Given that these
assumptions are met, our method is new exploratory tool that can help inform
more rigorous studies of parasite-induced host mortality.**

2. Pg 8-9 - to me it could be made clearer why we should care about the Type I error - it is just stated on Pg 9 with no justification. I am not saying the authors are incorrect. I would get this justification in at the start of the section - ie lines 150~ Pg 8. In addition, I would insert here the "by estimating power"after "pre-mortality parameters" (Line 154) and then expand a bit.  Just needs for me a couple of sentences setting up Type 1 and power and why done for the reader.  The more detailed description that follows then flows.

To address these comments, we added the following justification for using Type I error and power at the beginning the subsection "Evaluating the Adjei and Lieklihood Methods"

**We used statistical power and Type I error to test the ability of the Adjei and the Likelihood Methods to correctly identify the presence of PIHM on simulated data with known pre-mortality parameters. The power of a method is the probability of correctly detecting PIHM given that it is occurring and the Type I error is the probability of incorrectly identifying PIHM given that it is not occurring. If a method has low Type I error we can be confident that when we detect PIHM it is actually occurring.  If one method has higher power for detecting PIHM than another, we will need to sample fewer hosts to detect PIHM.**

3. Pg 8 Line 169-175 - this may be V2 of your work and not for this paper, but I wonder what would be the impact of non-random mortality. I am thinking here of say where mortality is a combination of burden and some host factor - eg sex. As said this maybe beyond the scope of the paper, but could be mentioned in the discussion if agree?

This is a great point by Reviewer 2.

4. lines 201-202 - I assume that any confidence intervals around the average no of surviving hosts are nice and symmetrical - ie that we are happy to just take the mean and do not need to worry about some horrible non linear type distribution of surviving hosts.

Reviewer 2 is correct in assuming that the distribution of surviving hosts was generally quite symmetrical and the standard deviation was small relative to the mean. We made this clear in the manuscript by adding the following line

**The distribution of surviving hosts over the 150 simulations was generally symmetrical and the standard deviation was small compared to the mean (maximum coefficient of variation was approximately 0.06 across all parameter combinations), suggesting that the mean number of surviving hosts was an adequate summary statistic of the number of hosts sampled post-mortality.**

5. Pg 13 - lines 295-302 - is it of interest that the variation your ML method observed in the LD50 seemed a lot less than the Adjei method?  Given this is the same host, same parasite and similar habitats I would be a little concerned at the marked variation in LD50 in the Adjei method

6. Pg 13-14 - lines 316-> - see comments above about being more positive about your work; Pg 15 - I think the conclusions need to be more positive

We addressed these concerns in above section.

7. Figure 2, 3, supplementary figures - distinction between ML and Adeji not always very clear.  Would suggest (a) reordering the legend to ML grad, mode,steep then the Adjei equivalents; (b) having the lines in the legend box longer; (c) ideally some colour to distinguish between ML and Adjei, if not put into the figure legends themselves Adjei=gray, ML=black; (d) can the lines be any thicker on the graphs?

We changed Figures 2, 3 and corresponding supplementary figures according the the suggestions by the reviewer. We reordered the legends, lengthened the lines in the legend box, colored the Adjei Lines green, and made all the lines thicker.  We believe that this makes the figures much more readable.

8. Figure 4 - I would recommend having vertical lines coming down from the steep and moderate to see what sample size is associated with 80% power

Added the recommended vertical lines to the Figure.

9. Figure S4 missing from my download

Seems to be present in the version that we are looking at.

10. Pg 2 lines 13-15 ref for sentence would be good

Added a citation AndersonandMay1978

11. Pg 5 line 84 - an unnecessary "an"

Deleted the extra "an"

12. Pg 12 - Figure 4 referred to in text before Figure 3 - suggest swapping order of figures

Swapped the order of the figures

### 2015-08-08 docs/parasite_mortality_manuscript.tex
Incorporated reviewer comments

Infectious agents can impact animal populations by changing
population dynamics and stability \citep{Dobson1992,Tompkins2002}, altering predator-prey interactions \citep{Joly2004}, and
even causing species' decline and extinction \citep{DeCastro2005a,McCallum2012b}. Accurately estimating the impact
of these infectious agents in wildlife is critical to understanding what
regulates host and parasite populations, making predictions about disease
transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies \citep{Coyne1989}, bovine tuberculosis \citep{Cox2005}, and
rinderpest \citep{Tille1991}, are typically modeled based on the presence or absence of disease, such that host survival is not typically considered to be a function of the number of infectious agents present within the host.  In contrast, models of macroparasites generally assume that pathology increases with parasite burden and host survival probability must be treated as a function of infection intensity. Helminths exhibiting this intensity-dependent pathology have significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013}, and wildlife survival \citep{Kirk2003,Logiudice2003}. While it is generally assumed that some fraction of wild host populations succumb to parasitic infection, it is notoriously difficult to actually quantify parasite-induced host mortality (PIHM) in wild animal populations because it is difficult to observe the dead or dying hosts most impacted by parasitism \citep{McCallum2000a}.

@@

Our method (henceforth the Likelihood Method) begins with the same assumptions as the Adjei Method: namely that infection has occurred and hosts with fatal parasite loads have died prior to the population sampling. As discussed by
\citeauthor{Adjei1986}, this is not necessarily an unrealistic as some parasite infections occur primarily in younger hosts with parasite-induced mortality occurring soon after infection \citep[e.g.][]{Schotthoefer2003,Johnson2008}.

@@

We tested the ability of the Adjei and the Likelihood Methods to identify the presence of PIHM on simulated data with known pre-mortality parameters. Consistent with the model assumption that parasite infection, host mortality, and population sampling are temporally separate events, we first created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
negative binomial distribution with parameters $\mu_p$ and $k_p$. This represents a host population that has become infected but not yet experienced parasite-induced mortality \citep{Adjei1986}.  In the Adjei Method and Crofton Method, $N_p$ is a necessary parameter defined as the number of hosts in the population before parasite-induced mortality. More accurately, $N_p$  is the number of hosts that would have been sampled had parasite-induced host mortality not occurred.  This parameter is not necessary when using the Likelihood Method because, unlike the Adjei Method and Crofton Method which estimate parasite-induced mortality using absolute numbers of hosts, the Likelihood Method estimates parasite-induced mortality using probabilities. However, to compare the results of the Likelihood Method with the Adjei Method, we specified a value for $N_p$ for all simulations.

@@

We compared the two methods using three different mean parasite intensity values ($\mu_p$ = 10, 50, 100) and three different host survival functions (gradual, moderate, and steep decreases in the host survival with increasing parasite intensity, Figure \ref{fig:question1}A). For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.  We examined each $\mu_p$-survival function pair at  three levels of parasite
aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these 27 parameter
combinations we simulated 150 datasets and tested the probability of each method correctly identifying PIHM in the post-mortality dataset (power) and incorrectly identifying PIHM in the pre-mortality dataset (Type I error).  For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level of 0.05.  We also examined the impact of sample size by simulating each parameter for pre-mortality sample sizes of $N_p$ = [50, 100, 200, 300, 400, 500].  Wild host populations were assumed to be sampled after PIHM has occurred, thus we calculated the sample size in the power simulations as the average number of surviving hosts over all 150 simulations for each parameter combination.

@@

While we have improved on the existing methods for quantifying
PIHM from parasite intensity data, all such methods require several
fundamental, and potentially problematic assumptions.  Nearly all current methods derive from \cite{Crofton1971a} \citep[but see][]{Ferguson2011} and assume that, prior to any PIHM, parasites are distributed in the host population following a
negative binomial distribution. But, it is fundamentally impossible to know
what the pre-mortality parasite distribution was in a wild host population and
it is widely recognized that different processes can lead to a variety of
parasite distributions in hosts \citep{Anderson1982a, Duerr2003}. However, the negative binomial is extremely
flexible and there is substantial empirical and theoretical evidence to support
the assumption that, prior to any PIHM, parasite distributions can be fit by a negative binomial distribution \citep{Shaw1995,Shaw1998,Wilson2002}.

Unfortunately, this
flexibility in the distribution may also reduce our ability to detect PIHM. If
a negative binomial can be fit to the observed post-mortality parasite
distribution then, regardless of how lethal the parasite was, it will be
impossible to detect PIHM because there is no need for a more complex model.
Most observed parasite distributions are well fit by the negative binomial distribution \citep{Shaw1998}, suggesting
that systems where these methods are applicable may be more the exception than
the rule.  Furthermore, even when truncation of the negative binomial distribution is detected, it may be caused by other
processes such as within host density dependence, age dependent variation in host
resistance and/or heterogeneous infection rates \citep{Anderson1982a,Rousset1996, McCallum2000a}.  This means that in the event
that PIHM is detected, it may actually not be the result of PIHM.

Given these numerous caveats, is there a place in parasitology for methods that
estimate PIHM from intensity data?  As suggested by
\cite{Lester1984} these methods for estimating PIHM can
provide preliminary insight into whether or not PIHM is worth further
exploration.  However, we stress that these methods should only be used as an
exploratory tool when assessing the role of PIHM in a system, and potential
users should critically evaluate whether they think they have a large enough
sample size and an appropriate host survival function/post-mortality distribution for the methods developed
in this paper to be applicable.  Even if they are applicable, inferring PIHM
from distributional data is no substitute for field experiments
and an in depth understanding of the natural history of the host-parasite
system under consideration.

@@

@@

    \caption{The power of the Likelihood Method (Like.) to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated. The curves were generated from 500 simulations for 10 pre-mortality sample sizes, $N_p$.}

>>>>

Infectious agents can impact animal populations by changing
population dynamics and stability \citep{Dobson1992,Tompkins2002}, altering predator-prey interactions \citep{Joly2004}, and
even causing species' decline and extinction \citep{DeCastro2005a,McCallum2012b}. Accurately estimating the impact
of these infectious agents in wildlife is critical to understanding what
regulates host and parasite populations, making predictions about disease
transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies \citep{Coyne1989}, bovine tuberculosis \citep{Cox2005}, and
rinderpest \citep{Tille1991}, are typically modeled based on the presence or absence of disease, such that host survival is not typically considered to be a function of the number of infectious agents present within the host.  In contrast, models of macroparasites generally assume that pathology increases with parasite burden and host survival probability must be treated as a function of infection intensity \citep{AndersonandMay1978}. Helminths exhibiting this intensity-dependent pathology have significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013}, and wildlife survival \citep{Kirk2003,Logiudice2003}. While it is generally assumed that some fraction of wild host populations succumb to parasitic infection, it is notoriously difficult to actually quantify parasite-induced host mortality (PIHM) in wild animal populations because it is difficult to observe the dead or dying hosts most impacted by parasitism \citep{McCallum2000a}.

@@

Our method (henceforth the Likelihood Method) begins with the same assumptions as the Adjei Method: namely that infection has occurred and hosts with fatal parasite loads have died prior to the population sampling. As discussed by
\citeauthor{Adjei1986}, this is not necessarily unrealistic as some parasite infections occur primarily in younger hosts with parasite-induced mortality occurring soon after infection \citep[e.g.][]{Schotthoefer2003,Johnson2008}.

@@

We used statistical power and Type I error to test the ability of the Adjei Method and the Likelihood Method to correctly identify the presence of PIHM on simulated data with known pre-mortality parameters. The power of a method is the probability of correctly detecting PIHM given that it is occurring and the Type I error is the probability of incorrectly identifying PIHM given that it is not occurring. If a method has low Type I error we can be confident that when we detect PIHM it is actually occurring.  If one method has higher power for detecting PIHM than another, we will need to sample fewer hosts to detect PIHM.

Consistent with the model assumption that parasite infection, host mortality, and population sampling are temporally separate events, we first created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
negative binomial distribution with parameters $\mu_p$ and $k_p$. This represents a host population that has become infected but not yet experienced parasite-induced mortality \citep{Adjei1986}.  In the Adjei Method and Crofton Method, $N_p$ is a necessary parameter defined as the number of hosts in the population before parasite-induced mortality. More accurately, $N_p$  is the number of hosts that would have been sampled had parasite-induced host mortality not occurred.  This parameter is not necessary when using the Likelihood Method because, unlike the Adjei Method and Crofton Method which estimate parasite-induced mortality using absolute numbers of hosts, the Likelihood Method estimates parasite-induced mortality using probabilities. However, to compare the results of the Likelihood Method with the Adjei Method, we specified a value for $N_p$ for all simulations.

@@

We compared the two methods using three different mean parasite intensity values ($\mu_p$ = 10, 50, 100) and three different host survival functions (gradual, moderate, and steep decreases in the host survival with increasing parasite intensity, Figure \ref{fig:question1}A). For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.  We examined each $\mu_p$-survival function pair at  three levels of parasite
aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these 27 parameter
combinations we simulated 150 datasets and tested the probability of each method correctly identifying PIHM in the post-mortality dataset (power) and incorrectly identifying PIHM in the pre-mortality dataset (Type I error).  For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level of 0.05.  We also examined the impact of sample size by simulating each parameter for pre-mortality sample sizes of $N_p$ = [50, 100, 200, 300, 400, 500].  Wild host populations were assumed to be sampled after PIHM has occurred, thus we calculated the sample size in the power simulations as the average number of surviving hosts over all 150 simulations for each parameter combination. The distribution of surviving hosts over the 150 simulations was generally symmetrical and the standard deviation was small compared to the mean (maximum coefficient of variation was approximately 0.06 across all parameter combinations), suggesting that the mean number of surviving hosts was an adequate summary statistic of the number of hosts sampled post-mortality.

@@

While we have improved on the existing methods for quantifying
PIHM from parasite intensity data, all such methods require several
fundamental assumptions.  Nearly all current methods derive from \cite{Crofton1971a} \citep[but see][]{Ferguson2011} and assume that, prior to any PIHM, parasites are distributed in the host population following a
negative binomial distribution. But, it is fundamentally impossible to know
what the pre-mortality parasite distribution was in a wild host population and
it is widely recognized that different processes can lead to a variety of
parasite distributions in hosts \citep{Anderson1982a, Duerr2003}. However, the negative binomial is extremely
flexible and there is substantial empirical and theoretical evidence to support
the assumption that, prior to any PIHM, parasite distributions can be fit by a negative binomial distribution \citep{Shaw1995,Shaw1998,Wilson2002}.

It is important to note that the flexibility of the negative binomial distribution may also reduce our ability to detect PIHM. If
a negative binomial can be fit to the observed post-mortality parasite
distribution then, regardless of how lethal the parasite was, it will be
impossible to detect PIHM because there is no need for a more complex model.
Many observed parasite distributions are well-fit by the negative binomial distribution \citep{Shaw1998}, suggesting
that systems where these methods are applicable without any a prior knowledge may be uncommon. However, if one has a prior knowledge about some aspect of the pre-mortality distribution \citep[e.g. assumes/knows the value of $k_p$,][]{Ferguson2011}, then the Likelihood Method could be applicable even if the the post-mortality distribution was well-fit by a negative binomial.

If one has evidence that the pre-mortality is not negative binomial, the generality of our method easily allows another distribution to be specified for $g(x, \boldsymbol{\phi})$. In particular, one could use the
resulting stationary host-parasite distribution from a stochastic host-parasite
model without parasite-induced host mortality \citep{Anderson1982a} to specify
the form of $g(x, \boldsymbol{\phi})$ and then apply the techniques discussed
in this paper to detect PIHM.  The general criteria necessary
for the Likelihood Method to detect PIHM in a stochastic host-parasite process
is that the stationary distribution of the process with
mortality is significantly different than the stationary distribution without mortality.
It is widely recognized that parasite-induced host mortality decreases the
aggregation of host-parasite distributions relative to those without mortality
\citep{Barbour2000}, suggesting that Likelihood Method could be applicable for
detecting PIHM in host-parasite systems experiencing stochastic parasite births, deaths, and immigrations in addition to host mortality.

If the Likelihood Method is applicable and the truncation of the negative binomial distribution is detected, one must be aware that the pattern may be caused by other
processes such as within host density dependence, age dependent variation in host
resistance and/or heterogeneous infection rates \citep{Anderson1982a,Rousset1996, McCallum2000a}. This means that in the event
that PIHM is detected, it may actually not be the result of PIHM.  Moreover, if host mortality depends on parasite intensity and additional variables (e.g. host sex, host size), failure to identify these important confounding variables could significantly affect the ability of these methods to correctly identify PIHM. However, both of these issues -- inferring process from pattern and confounding variables -- are well-recognized limitations of most statistical inference and are addressed via judicious model specification and selection \citep{Seber2003}.

As suggested by \cite{Lester1984} these methods for estimating PIHM can
provide preliminary insight into whether or not PIHM is worth further
exploration.  However, we stress that these methods are an
exploratory tool for assessing the role of PIHM in a system, and potential
users should critically evaluate whether they think they have a large enough
sample size and an appropriate host survival function/post-mortality distribution for the methods developed
in this paper to be applicable.  Even if they are applicable, inferring PIHM
from distributional data is no substitute for field experiments
and an in depth understanding of the natural history of the host-parasite
system under consideration.

@@

    \caption{The power of the Likelihood Method (Like.) to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated. The curves were generated from 500 simulations for 10 pre-mortality sample sizes, $N_p$. The vertical, dotted-dashed lines indicate the sample size at which the power the for the Likelihood Method with steep and moderate survival functions is 0.8 (75 hosts for steep functions and 402 for moderate functions). The Likelihood Method with a gradual survival function never has a power above 0.8.}

@@
